---
title: "Kknn_opt"
output: html_document
date: "2025-05-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(kknn)
library(dplyr)
library(ggplot2)
library(caret)
library(pROC)
library(smotefamily)
library(knitr)
```

# Modeling Strategy

**1. Logistic Regression**  
- Instance-based learner that adapts to local patterns and nonlinear decision boundaries using weighted distances.  
- Works well when there’s a linear relationship between features and the log-odds of the classes.  

**2. Kernel k-Nearest Neighbors (kknn)**  
- Simple and effective for multiclass classification. Captures local data structure and doesn’t assume any distribution.  
- Ideal for multiclass classification with no distributional assumptions, especially after balancing techniques like SMOTE.  

**3. Support Vector Machines (SVM)**  
- Performs well in high-dimensional spaces. Robust to outliers with the right kernel and effective on small- to medium-sized datasets.  
- Suitable when data is not linearly separable — kernels like RBF help in such cases.  

**4. Random Forest**  
- An ensemble method that reduces overfitting and handles both categorical and numerical features well.  
- Strong performance on noisy or imbalanced datasets and can handle many input variables without the need for variable deletion.  

**5. Gradient Boosted Trees (XGBoost)**  
- A state-of-the-art boosting technique with high performance and built-in regularization to prevent overfitting.  
- Handles missing values natively and excels in structured/tabular data, such as credit scoring datasets.


---
# Plan for Evaluation

The following metrics were used for model evaluation along with cross-validation:

**Accuracy:** Gives a quick view of overall correctness, useful when class distribution is fairly balanced.  
**Precision:** Helps reduce false positives — critical in credit scoring to avoid mistakenly approving risky profiles.  
**Recall:** Ensures we catch most of the truly risky credit cases, minimizing false negatives.  
**F1-Score:** Balances precision and recall, making it ideal when we care equally about both types of error, especially in imbalanced datasets.  
**ROC-AUC:** Measures the model’s ability to separate classes across all thresholds. AUC near 1 means strong overall classification performance.  
**Cross-Validation:** We use k-fold cross-validation to ensure our model’s performance is *reliable and generalizes well* to unseen data, avoiding overfitting or underfitting due to random train-test splits.

**We chose ROC-AUC as the primary metric for determing the best model because it gives a comprehensive evaluation of the model's ability to distinguish between classes, especially important in our imbalanced, multiclass credit scoring problem. It helps ensure reliable decision-making beyond raw accuracy.**

</div>

---
# Kernel K-Nearest Neighbours 

```{r}
# Load data
df <- read.csv("../data/clean_train.csv")
df$Credit_Score <- as.factor(df$Credit_Score)

```

### Modelling Process
The kknn model is trained on a SMOTE-balanced dataset to address class imbalance.

It uses the train.kknn() function to automatically perform internal cross-validation over a range of k values (1–21), and multiple kernel functions (rectangular, epanechnikov, triangular, optimal).

For each kernel, the best k is selected based on lowest misclassification rate (i.e., highest accuracy).

The model with triangular kernel and k=5 gave the best balance of accuracy and AUC and was chosen for final evaluation.

## Data Preprocessing

```{r}
# Drop high-cardinality categorical columns
df_knn <- df %>%
  select(-Customer_ID, -Month, -Occupation, -Credit_Mix, 
         -Payment_of_Min_Amount, -Payment_Behaviour)

```

```{r}
set.seed(5003)

# Separate predictors and response
X <- df_knn[, -ncol(df_knn)]
y <- df_knn$Credit_Score

# Apply SMOTE
smote_result <- SMOTE(X, y, K = 5)

# Preparing balanced data
df_balanced <- smote_result$data
df_balanced$Credit_Score <- as.factor(df_balanced$class)
df_balanced$class <- NULL  # Remove redundant column

```

```{r}
# Split data into training and test sets
set.seed(5003)
split_idx <- createDataPartition(df_balanced$Credit_Score, p = 0.8, list = FALSE)
train_kknn <- df_balanced[split_idx, ]
test_kknn <- df_balanced[-split_idx, ]

```

```{r}
# Scale numerical data
preproc <- preProcess(train_kknn[, -ncol(train_kknn)], method = c("center", "scale"))
train_scaled <- predict(preproc, train_kknn[, -ncol(train_kknn)])
test_scaled <- predict(preproc, test_kknn[, -ncol(test_kknn)])

train_scaled$Credit_Score <- train_kknn$Credit_Score
test_scaled$Credit_Score <- test_kknn$Credit_Score

```

## Model Training to find best Model

```{r}
# Try multiple kernels
kernels <- c("rectangular", "epanechnikov", "triangular", "optimal")
results_list <- list()
best_model <- NULL
best_metric <- 0 
best_kernel <- NULL

start <- Sys.time()

for (kernel in kernels) {
  cat("\nTraining with kernel:", kernel, "\n")
  
  # Train model
  set.seed(5003)
  model <- train.kknn(Credit_Score ~ ., data = train_scaled,
                      ks = 5:15, kernel = kernel, distance = 2)
  
  # Predict on test set
  pred <- predict(model, test_scaled)
  cm <- confusionMatrix(pred, test_scaled$Credit_Score)
  auc <- multiclass.roc(test_scaled$Credit_Score, as.numeric(pred))$auc
  
  # Macro-averaged metrics
  precision <- cm$byClass[, "Precision"]
  recall <- cm$byClass[, "Recall"]
  f1 <- cm$byClass[, "F1"]
  
  acc <- round(cm$overall["Accuracy"], 4)
  
  results_list[[kernel]] <- data.frame(
    Kernel = kernel,
    Best_K = model$best.parameters$k,
    Accuracy = acc,
    Kappa = round(cm$overall["Kappa"], 4),
    Macro_Precision = round(mean(precision, na.rm = TRUE), 4),
    Macro_Recall = round(mean(recall, na.rm = TRUE), 4),
    Macro_F1 = round(mean(f1, na.rm = TRUE), 4),
    AUC = round(auc, 4)
  )
  
  # Check if current model is best so far
  if (acc > best_metric) {
    best_metric <- auc
    best_model <- model
    best_kernel <- kernel
  }
}

end <- Sys.time()
print(paste("Total optimization time:", round(end - start, 2)))

# Save the best model
saveRDS(best_model, file = "best_kknn_model.rds")
cat("Best model saved! Kernel:", best_kernel, "Accuracy:", best_metric, "\n")


```

```{r}
# Combine all results and display as a table
optim_results <- do.call(rbind, results_list)
knitr::kable(optim_results, caption = "Optimized KNN Results with Different Kernels")

```

```{r}
saveRDS(model, "kknn_opt.rds")
```

## Using the best Model 
### KNN with kernel 'Triangular' with k values ranging from 1-21

```{r}
set.seed(5003)
kknn_model <- train.kknn(Credit_Score ~ ., data = train_scaled,
                         ks = 1:21, kernel = "triangular", distance = 2)
```

```{r}
best_knn <- readRDS("kknn_model.rds")
```

### KKNN - Model Performance

#### training accuracy
```{r}
train_pred <- predict(best_knn, train_scaled)
train_cm <- confusionMatrix(train_pred, train_scaled$Credit_Score)
train_accuracy <- train_cm$overall["Accuracy"]
print(paste("Train Accuracy:", round(train_accuracy, 4)))

```
#### test accuracy
```{r}
test_pred <- predict(best_knn, test_scaled)
test_cm <- confusionMatrix(test_pred, test_scaled$Credit_Score)
test_accuracy <- test_cm$overall["Accuracy"]
print(paste("Test Accuracy:", round(test_accuracy, 4)))

```
```{r}
# Confusion Matrix
confusionMatrix(test_pred, test_scaled$Credit_Score)
```
- **Test Accuracy:** 0.83196 – strong performance on unseen data.  
- **Training Accuracy:** 0.982 – very high, but test accuracy confirms minimal overfitting.  
- **Precision, Recall, F1-Score:** All approximately 0.80–0.81 – indicates balanced prediction across all credit classes.  
- **AUC-ROC:** 0.8408 – demonstrates strong ability to distinguish between classes.


## Visualisation

### Best k vs Accuracy Plot

```{r}
plot(best_knn)
```

```{r}
# Extract misclassification as numeric vector
misclass_vec <- as.numeric(best_knn$MISCLASS)

# Create accuracy dataframe
plot_data <- data.frame(
  K = 1:length(misclass_vec),
  Misclassification = misclass_vec,
  Accuracy = 1 - misclass_vec
)

# Plot Accuracy vs K
library(ggplot2)
ggplot(plot_data, aes(x = K, y = Accuracy)) +
  geom_line(color = "blue") +
  geom_point(color = "red", size = 2) +
  labs(title = "Accuracy vs K (Triangular Kernel)", x = "K", y = "Accuracy") +
  theme_minimal()

```

- Peak model accuracy observed at k = 5.
- Accuracy declines after k=5, indicating loss of local sensitivity with larger k.

### Confusion Matrix
```{r}
cm_table <- as.data.frame(test_cm$table)

ggplot(cm_table, aes(Prediction, Reference, fill = Freq)) +
  geom_tile() +
  geom_text(aes(label = Freq), color = "white", size = 5) +
  scale_fill_gradient(low = "lightblue", high = "darkblue") +
  labs(title = "Confusion Matrix (Test Set - Triangular Kernel)", x = "Predicted", y = "Actual") +
  theme_minimal()

```

### ROC Curve

```{r}
# Get predicted probabilities for each class
prob_pred <- predict(best_knn, test_scaled, type = "prob")

par(mfrow = c(1, 1))

# Get actual labels
actual <- test_scaled$Credit_Score

# Create ROC for each class
classes <- levels(actual)
colors <- c("red", "blue", "darkgreen")
for (i in seq_along(classes)) {
  # Binary label for this class
  bin_actual <- as.numeric(actual == classes[i])
  
  # ROC curve for this class
  roc_i <- roc(bin_actual, prob_pred[, i])
  
  # Plot
  if (i == 1) {
    plot(roc_i, col = colors[i], main = "One-vs-Rest ROC Curves (Triangular Kernel)", lwd = 2)
  } else {
    lines(roc_i, col = colors[i], lwd = 2)
  }
}

legend("bottomright", legend = paste("Class", classes), col = colors, lwd = 2)
```
- "Good" credit class is detected very well (high sensitivity and precision).
- "Poor" and "Standard" classes are moderately detected — model struggles more with differentiating these.

## Final Summary

```{r}
# Accuracy, precision, recall, F1
accuracy <- test_cm$overall["Accuracy"]
precision <- mean(test_cm$byClass[, "Precision"], na.rm = TRUE)
recall <- mean(test_cm$byClass[, "Recall"], na.rm = TRUE)
f1 <- mean(test_cm$byClass[, "F1"], na.rm = TRUE)

# AUC using probabilities (only do this once)
auc <- multiclass.roc(test_scaled$Credit_Score, as.numeric(test_pred))$auc

# Create metrics table
metrics_df <- data.frame(
  Metric = c("Accuracy", "Precision", "Recall", "F1-Score", "AUC-ROC"),
  Value = round(c(accuracy, precision, recall, f1, auc), 6)
)

# Display
kable(metrics_df, caption = "Model Performance Metrics (Triangular Kernel)")

```

- The kknn model (triangular kernel) offers excellent predictive power, particularly for identifying "Good" credit scores, with robust generalization and minimal signs of overfitting.